Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-20 reproducibility/manuscript/manuscript.qmd
docs(manuscript): set fig prefix `Fig.`
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: "Pyro-Velocity: Probabilistic RNA Velocity inference from single-cell data"
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    classoption: 
      - lineno
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: citeproc
    keep-tex: true
    equal-margins: false
    include-in-header: header.tex
  html:
    code-fold: true
    html-math-method: mathjax
  docx:
    reference-doc: styles.docx
# filters:
#   - authors-block
csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
affiliations:
  - name: Massachusetts General Hospital Research Institute, Charlestown, MA
    id: 1
    department: Molecular Pathology Unit
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Harvard Medical School, Charlestown, MA
    id: 2
    department: Massachusetts General Hospital Cancer Center
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Massachusetts General Hospital, Boston, MA
    id: 3
    department: Center for Regenerative Medicine
    city: Boston
    state: MA
    postal-code: 02114
  - name: Harvard University, Cambridge, MA
    id: 4
    department: Harvard Stem Cell Institute
    city: Cambridge
    state: MA
    postal-code: 02139
  - name: Cambridge, MA
    id: 5
    department: Broad Institute of MIT and Harvard
    city: Cambridge
    state: MA
    postal-code: 02142
  - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
    id: 6
    department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
    city: Lausanne
    state: Switzerland
    postal-code: 1015
  - name: Cambridge, MA
    id: 7
    department: Basis Research Institute
    city: Cambridge
    state: MA
    postal-code: 02142
authors:
  - name: Qian Qin
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
      - ref: 5
    # attributes:
    #   equal-contributor: true
  - name: Eli Bingham
    affiliations:
      - ref: 5
      - ref: 7
  - name: Gioele La Manno
    affiliations:
      - ref: 6
    email: gioele.lamanno@epfl.ch
    attributes:
      corresponding: true
  - name: David M. Langenau
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
    email: dlangenau@mgh.harvard.edu
    attributes:
      corresponding: true
  - name: Luca Pinello
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 5
    email: lpinello@mgh.harvard.edu
    attributes:
      corresponding: true
abstract: |
  Single-cell RNA Velocity has dramatically advanced our ability to model cellular differentiation and cell fate decisions. However, current preprocessing choices and model assumptions often lead to errors in assigning developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity model to estimate the uncertainty of cell future states. This approach models raw sequencing counts with the synchronized cell time across all expressed genes to provide quantifiable and improved information on cell fate choices and developmental trajectory dynamics.
keywords: [RNA velocity, single-cell RNA sequencing, probabilistic generative modeling, cell fate, developmental trajectory inference]
bibliography: bibliography.bib
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

@@

RNA Velocity is a powerful computational framework to estimate the time derivative of gene expression [@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj]. The framework has been used to study developmental cell lineage trajectories over a broad spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj], pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors differentiation [@Bergen2020-pj]. Yet, recent state-of-the-art RNA Velocity approaches such as velocyto and scVelo [@La_Manno2018-lj] have several limitations. First, they may predict developmental trajectories that do not exist in nature, as it has been shown in the analysis of fully mature blood cells [@Zheng2017-bz;@Bergen2021-qz]. Second, data preprocessing steps (e.g., selection of top variable genes, dimensionality reduction) affect velocity estimates that are difficult to characterize and might lead to artefactual predictions. In addition, non-linear spliced and unspliced imputation procedures could distort phase portrait geometry and recovered trajectories (Fig. 1a). Third, most RNA velocity algorithms perform estimates independently for each gene, and therefore cannot exploit the natural interdependence of gene expression. Fourth, a unified description of the temporal relation between cells can only be assigned based on postprocessing steps that aggregate heuristically the gene-wise fits into a shared latent time (Fig. 1a). Finally, current RNA velocity frameworks only output point estimates of the velocity vector field and do not provide uncertainty estimation. Importantly, the absence of statistical confidence estimates increases false discoveries as it leaves users without a way to evaluate the robustness of the methods’ outputs. To overcome these limitations, we developed a probabilistic generative model, named Pyro-Velocity, that estimates RNA dynamics and infers cell trajectories using Pyro8, a robust and efficient probabilistic programming language.

@fig-conceptual-model. @fig-model-results. @suppfig-graphical-models.

@@

@@

\FloatBarrier
\newpage

@@

![Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation of velocity vector fields and shared latent time, based on raw counts and without ad-hoc preprocessing steps. a. Schematic comparing current RNA Velocity and Pyro-Velocity workflows. Both methods take in input spliced and unspliced read count tables and output RNA velocity vector fields and latent (shared) time. Previous methods require ad-hoc preprocessing steps, including the arbitrary selection of top variable genes, number of principal components, and kNN-pooling to impute both spliced and unspliced expression within the PCA-reduced spliced expression space. In addition, manually tuned ad-hoc postprocessing steps are used to compute the final vector field and latent time based on aggregation and smoothing of the independent velocity estimates for each gene which can dramatically change the velocity estimates. In contrast, Pyro-Velocity is a fully generative and probabilistic model that uses raw read counts. Our model requires minimal preprocessing and postprocessing steps and outputs a common shared time based on all the genes without additional aggregation steps. It also provides uncertainty estimation based on posterior samples. b. The main outputs of the Pyro-Velocity framework. Visualization of single gene expression across inferred developmental cell order using rainbow plots (right panel).](figures/fig.pdf){#fig-conceptual-model}

>>>>

---
title: "Pyro&thinsp;-Velocity: Probabilistic RNA Velocity inference from single-cell data"
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    classoption: 
      - lineno
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: citeproc
    keep-tex: true
    equal-margins: false
    include-in-header: header.tex
  html:
    code-fold: true
    html-math-method: mathjax
  docx:
    reference-doc: styles.docx
# filters:
#   - authors-block
csl: _extensions/sciexp/nature/csl/springer-basic-brackets-no-et-al.csl
affiliations:
  - name: Massachusetts General Hospital Research Institute, Charlestown, MA
    id: 1
    department: Molecular Pathology Unit
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Harvard Medical School, Charlestown, MA
    id: 2
    department: Massachusetts General Hospital Cancer Center
    city: Charlestown
    state: MA
    postal-code: 02129
  - name: Massachusetts General Hospital, Boston, MA
    id: 3
    department: Center for Regenerative Medicine
    city: Boston
    state: MA
    postal-code: 02114
  - name: Harvard University, Cambridge, MA
    id: 4
    department: Harvard Stem Cell Institute
    city: Cambridge
    state: MA
    postal-code: 02139
  - name: Cambridge, MA
    id: 5
    department: Broad Institute of MIT and Harvard
    city: Cambridge
    state: MA
    postal-code: 02142
  - name: Laboratory of Neurodevelopmental Systems Biology, Lausanne, Switzerland
    id: 6
    department: Brain and Mind Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL)
    city: Lausanne
    state: Switzerland
    postal-code: 1015
  - name: Cambridge, MA
    id: 7
    department: Basis Research Institute
    city: Cambridge
    state: MA
    postal-code: 02142
authors:
  - name: Qian Qin
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
      - ref: 5
    # attributes:
    #   equal-contributor: true
  - name: Eli Bingham
    affiliations:
      - ref: 5
      - ref: 7
  - name: Gioele La Manno
    affiliations:
      - ref: 6
    email: gioele.lamanno@epfl.ch
    attributes:
      corresponding: true
  - name: David M. Langenau
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
      - ref: 4
    email: dlangenau@mgh.harvard.edu
    attributes:
      corresponding: true
  - name: Luca Pinello
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 5
    email: lpinello@mgh.harvard.edu
    attributes:
      corresponding: true
abstract: |
  Single-cell RNA Velocity has dramatically advanced our ability to model cellular differentiation and cell fate decisions. However, current preprocessing choices and model assumptions often lead to errors in assigning developmental trajectories. Here, we develop, Pyro-Velocity, a Bayesian, generative, and multivariate RNA Velocity model to estimate the uncertainty of cell future states. This approach models raw sequencing counts with the synchronized cell time across all expressed genes to provide quantifiable and improved information on cell fate choices and developmental trajectory dynamics.
keywords: [RNA velocity, single-cell RNA sequencing, probabilistic generative modeling, cell fate, developmental trajectory inference]
bibliography: bibliography.bib
crossref:
  fig-prefix: Fig.
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Fig. S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

@@

RNA Velocity is a powerful computational framework to estimate the time derivative of gene expression [@La_Manno2018-lj;@Svensson2018-vk;@Qiu2022-dj;@Bergen2020-pj]. The framework has been used to study developmental cell lineage trajectories over a broad spectrum of biological processes, such as  haematopoiesis [@Qiu2022-dj], pancreas development [@Bastidas-Ponce2019-lf], and neural progenitors differentiation [@Bergen2020-pj]. Yet, recent state-of-the-art RNA Velocity approaches such as velocyto and scVelo [@La_Manno2018-lj] have several limitations. First, they may predict developmental trajectories that do not exist in nature, as it has been shown in the analysis of fully mature blood cells [@Zheng2017-bz;@Bergen2021-qz]. Second, data preprocessing steps (e.g., selection of top variable genes, dimensionality reduction) affect velocity estimates that are difficult to characterize and might lead to artifactual predictions. In addition, non-linear spliced and unspliced imputation procedures could distort phase portrait geometry and recovered trajectories ([@fig-model]a). Third, most RNA velocity algorithms perform estimates independently for each gene, and therefore cannot exploit the natural interdependence of gene expression. Fourth, a unified description of the temporal relation between cells can only be assigned based on postprocessing steps that aggregate heuristically the gene-wise fits into a shared latent time ([@fig-model]a). Finally, current RNA velocity frameworks only output point estimates of the velocity vector field and do not provide uncertainty estimation. Importantly, the absence of statistical confidence estimates increases false discoveries as it leaves users without a way to evaluate the robustness of the methods’ outputs. To overcome these limitations, we developed a probabilistic generative model, named *Pyro&thinsp;-Velocity*, that estimates RNA dynamics and infers cell trajectories using Pyro [@Bingham2018-id], a robust and efficient probabilistic programming language.

@@

{{< pagebreak >}}

@@

\FloatBarrier
{{< pagebreak >}}

@@

::: {#fig-model}

![](figures/fig.pdf)

\label{fig-model-inputs} 
\label{fig-model-outputs}

Pyro-Velocity is a fully generative Bayesian method with uncertainty estimation of velocity vector fields and shared latent time, based on raw counts and without ad-hoc preprocessing steps. a. Schematic comparing current RNA Velocity and Pyro-Velocity workflows. Both methods take in input spliced and unspliced read count tables and output RNA velocity vector fields and latent (shared) time. Previous methods require ad-hoc preprocessing steps, including the arbitrary selection of top variable genes, number of principal components, and kNN-pooling to impute both spliced and unspliced expression within the PCA-reduced spliced expression space. In addition, manually tuned ad-hoc postprocessing steps are used to compute the final vector field and latent time based on aggregation and smoothing of the independent velocity estimates for each gene which can dramatically change the velocity estimates. In contrast, Pyro-Velocity is a fully generative and probabilistic model that uses raw read counts. Our model requires minimal preprocessing and postprocessing steps and outputs a common shared time based on all the genes without additional aggregation steps. It also provides uncertainty estimation based on posterior samples. b. The main outputs of the Pyro-Velocity framework. Visualization of single gene expression across inferred developmental cell order using rainbow plots (right panel).
:::
